---
title: "R Notebook"
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    toc: yes
---

This notebook records the full data analysis of the GutAlive+Filtering method study

#1. Load package and datasets
```{r, message=FALSE, warning=FALSE}
# load libraries
library(tidyr)
library(stringr)
library(DESeq2)
library(data.table)
library(dendextend)
library(here)
library(impute)
library(pvclust)
library(gplots)
library(RColorBrewer)
library(cluster)
library(xml2)
library(MSstatsTMT)
library(pcaMethods)
library(plotly)
library(htmlwidgets)
library(gridExtra)
library(factoextra)
library(zoo)
library(pheatmap)
library(corrplot)
library(ggcorrplot)

# read in data files
proteinGroups <- read.table("data/proteinGroups.txt", sep="\t", header=TRUE)
evidence <- read.table("data/evidence.txt", sep="\t", header=TRUE)
annotation.mq <- read.table("metadata/msstatstmt_annotation_GAF_simple.csv", sep=",", header=TRUE)
experiment_meta <- read.table("metadata/experiment_meta_simple.txt", sep="\t", header=TRUE)

#Subset sample ids by experiments
sample.filter <- as.list(fread("metadata/sampleID_filtering_list.txt", header = F))
sample.gutAlive <- as.list(fread("metadata/sampleID_gutAlive_list.txt", header = F))
```

#2. Process data table and normalize using MSstatsTMT
```{r, message=FALSE, warning=FALSE}
require("DreamAI")
# combine tables for MSstatsTMT
input.mq <- MaxQtoMSstatsTMTFormat(evidence, proteinGroups, annotation.mq)

# Normalization using MSstatsTMT
quant.msstats.log <- proteinSummarization(input.mq,
                                      method="LogSum",
                                      global_norm=TRUE,
                                      reference_norm=TRUE,
                                      remove_norm_channel = TRUE,
                                      remove_empty_channel = TRUE)


#quant.msstats$id <- str_c(quant.msstats$BioReplicate)
quant.msstats.log$ProteinLevelData$id <- str_c(quant.msstats.log$ProteinLevelData$BioReplicate)

# quant.msstats.log.wide <- quant.msstats.log$ProteinLevelData %>% pivot_wider(id_cols = Protein, names_from = id, values_from = Abundance) 

#subset data according to sample ids
quant.msstats.log.filter <- filter(quant.msstats.log$ProteinLevelData, str_detect(quant.msstats.log$ProteinLevelData$id, paste(sample.filter[[1]], collapse="|")))
quant.msstats.log.gutAlive <- filter(quant.msstats.log$ProteinLevelData, str_detect(quant.msstats.log$ProteinLevelData$id, paste(sample.gutAlive[[1]], collapse="|")))

#make wide tables
log_data.filter <- quant.msstats.log.filter %>% pivot_wider(id_cols = Protein, names_from = id, values_from = Abundance) 
log_data.gutAlive <- quant.msstats.log.gutAlive %>% pivot_wider(id_cols = Protein, names_from = id, values_from = Abundance) 

#write.table(log_data, "prosessed/processed_logSum_proteins.txt", sep = "\t")

```

#3. Perform PCA analysis
PCA analysis is performed using NIPALS method which is non-linear, this method avoids the batch effects due to missing values.

##3.1 Filtering test dataset
```{r, message=FALSE, warning=FALSE}
# use nipals to do pca
resNipals  <- pca(t(log_data.filter %>% select(-Protein)), method="nipals", center=T, nPcs=10)
tmt_scores <- scores(resNipals)
pca_data <- merge(tmt_scores, experiment_meta, by.x = 0, by.y= "BioReplicate")

pc12 <- ggplot(pca_data, aes(PC1, PC2, colour= Individual)) +
  geom_point(alpha = 0.5, aes(text= Sample, shape = Condition, size = Culture))+
  scale_shape_manual(values = 15:20) + theme_bw()
ggplotly(pc12, tooltip = "text")

# ggsave("Filter_pc12.pdf", width = 6, height = 4)

pc23 <- ggplot(pca_data, aes(PC2, PC3, colour= Individual)) +
  geom_point(alpha = 0.5, aes(text= Sample, shape = Condition, size = Culture))+
  scale_shape_manual(values = 15:20)
ggplotly(pc23, tooltip = "text")

```

##3.2 GutAlive test dataset
```{r, message=FALSE, warning=FALSE}
# use nipals to do pca
resNipals.ga  <- pca(t(log_data.gutAlive %>% select(-Protein)), method="nipals", center=T, nPcs=10)
tmt_scores.ga <- scores(resNipals.ga)
pca_data.ga <- merge(tmt_scores.ga, experiment_meta, by.x = 0, by.y= "BioReplicate")

pc12.ga <- ggplot(pca_data.ga, aes(PC1, PC2, colour= Culture)) +
            geom_point(alpha = 0.6, aes(text= Sample, shape = Condition), size = 3)+
            scale_shape_manual(values = 12:20)+ theme_bw()
ggplotly(pc12.ga, tooltip = "text")

# ggsave("GutAlive_pc12.pdf", width = 4, height = 3)

pc23.ga<- ggplot(pca_data.ga, aes(PC2, PC3, colour= Culture)) +
           geom_point(alpha = 0.6, aes(text= Sample, shape = Condition), size = 3)+
           scale_shape_manual(values = 12:20)+ theme_bw()
ggplotly(pc23.ga, tooltip = "text")
# ggsave("GutAlive_pc23.pdf", width = 4, height = 3)

```

#4. Perform Clustering
Distance is calculated between sample pairs, this takes commonly shared features between each pair of samples for each calculation.

##4.1 Filtering test dataset
```{r, message=FALSE, warning=FALSE, fig.width=6, fig.height=12}
#replacing sample label by actual sample id
hc.filter.data <- as.data.frame(t(log_data.filter %>% select(-Protein)))
hc.filter.data <- merge(experiment_meta%>% select(BioReplicate|Sample),
                        hc.filter.data, by.x = "BioReplicate", by.y= 0)
rownames(hc.filter.data) <- hc.filter.data$Sample

#calculate distance
dist.filter <- dist((hc.filter.data %>% select(-BioReplicate|-Sample)), method = "euclidean")
# do hierarchical cluster
hc.filter <- hclust(dist.filter, method = "ward.D2")

fviz_dend(hc.filter, k=9, cex = 0.5, main = "Filtering test dataset",
                  k_colors = c('#e6194b', '#3cb44b', '#ffe119', '#4363d8', '#f58231',
                               '#911eb4', '#46f0f0', '#f032e6', '#bcf60c'),
                  color_labels_by_k = TRUE, rect = T,
                  repel = T, # color labels by groups
                  xlab = "Samples", ylab = "Distance", sub = "", horiz = T)

ggsave("Filter_cluster.pdf", width = 5, height = 9)


```

##4.2 GutAlive dataset
```{r, message=FALSE, warning=FALSE, fig.width=6, fig.height=12}
#replacing sample label by actual sample id
hc.filter.data.ga <- as.data.frame(t(log_data.gutAlive %>% select(-Protein)))
hc.filter.data.ga <- merge(experiment_meta%>% select(BioReplicate|Sample),
                        hc.filter.data.ga, by.x = "BioReplicate", by.y= 0)
rownames(hc.filter.data.ga) <- hc.filter.data.ga$Sample

#calculate distance
dist.filter.ga <- dist((hc.filter.data.ga %>% select(-BioReplicate|-Sample)), method = "euclidean")
# do hierarchical cluster
hc.filter.ga <- hclust(dist.filter.ga, method = "ward.D2")

fviz_dend(hc.filter.ga, k=3, cex = 0.5, main = "GutAlive test dataset",
                  k_colors = c('#e6194b', '#3cb44b', '#4363d8'),
                  color_labels_by_k = TRUE, rect = T,
                  repel = T, # color labels by groups
                  xlab = "Samples", ylab = "Distance", sub = "", horiz = T)

```

#5. Visualize correlation matrix heatmap
Similiar to the calculation of distance, here we take commonly shared features between each pair of samples for each calculation of correlation.

##5.1 Filtering test dataset
```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=12}
# Prepare the correct format for calculating correlation
cor.filter.data <- hc.filter.data %>% select(-BioReplicate) %>% select(-Sample)

# calculate correlation matrix
cor.filter.matrix <- round(cor(t(cor.filter.data),  method = "spearman", use = "pairwise.complete.obs"), 2)

# draw correlation plot
ggcorrplot(cor.filter.matrix, hc.order = TRUE) +
    theme(axis.text.x = element_text(size=8,  angle=90),
          axis.text.y = element_text(size=8)) +
    scale_fill_gradient2(limit = c(0.5,1), low = "blue",
                         high =  "red", mid = "white",
                         midpoint = 0.75)
```

##5.2 GutAlive test dataset
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=10}
# Prepare the correct format for calculating correlation
cor.filter.data.ga <- hc.filter.data.ga %>% select(-BioReplicate) %>% select(-Sample)

# calculate correlation matrix
cor.filter.matrix.ga <- round(cor(t(cor.filter.data.ga),  method = "spearman", use = "pairwise.complete.obs"), 2)

# draw correlation plot
ggcorrplot(cor.filter.matrix.ga, hc.order = TRUE) +
    theme(axis.text.x = element_text(size=8,  angle=90),
          axis.text.y = element_text(size=8)) +
    scale_fill_gradient2(limit = c(0.5,1), low = "blue",
                         high =  "red", mid = "white",
                         midpoint = 0.75)
```
